---
title: "Basic Trip Segmentation"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Trip segmentation for animal tracking data

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.2888341.svg)](https://doi.org/10.5281/zenodo.2888341) [![Github All Releases](https://img.shields.io/github/downloads/abfleishman/trakR/total.svg)]()

Cite as: 
Fleishman, A.B. and Orben, R.A. 2019. trakR: Basic Animal Tracking Data Analysis Tools, V0.0.5. GitHub repository. <https://github.com/abfleishman/trakR> DOI:10.5281/zenodo.2888341

```{r load packages,echo=TRUE,results='hide',message=FALSE}

library(ggplot2)
library(maps)
library(mapdata)
library(dplyr)
library(stringr)
library(lubridate)
library(argosfilter)
```

# Install trakR package (lots of issues but this trip segmentation works currently)
```{r install trakR}
# install.packages("devtools") # for installing packages from github
# devtools::install_github("abfleishman/trakR",upgrade = "ask") # install my package

library(trakR)
```

# Load tracks 
This should be one file for all the tracks or create a data.frame with all the tracks.
The function `MakeTrip` is set up for processing multiple animals at once
 
For this tutorial we will use a single bird in the included `tracks` dataset.
```{r load data}
data("tracks")

head(tracks)
```

The `MakeTrip` function requiers a column with distances to the colony for each point.  The `Dist2Colony` function will calculate these distances. It requires the `argosfilter` package to run.

```{r dist2col}
library(argosfilter)
tracks$Dist2Col<-trakR::Dist2Colony(tracks = tracks, 
                                    dataLat = "Latitude",
                                    dataLon = "Longitude",
                                    ColonyLat = 56.60329,
                                    ColonyLong = -169.6760)
```

# Split Trips
The `makeTrip` function finds the points where a animal moves some treshold distance `DistCutOff` away from the colony and the points where the animal returns within that threshold and annotates the dataframe with two new column, one indicating those "In" and "Out" points and another with trip numbers.  
```{r make trips}

tracks_w_trips<-trakR::MakeTrip(tracks = tracks,
                                ID = "CaptureID",
                                DistCutOff = 0.1,
                                Dist2Colony = "Dist2Col",
                                NumLocCut = 3)
head(tracks_w_trips)

# Plot a bird to check
ggplot(tracks_w_trips,aes(Longitude,Latitude,col=factor(TripNum)))+
  geom_path(size=.7)+
  geom_point(data=tracks_w_trips[tracks_w_trips$ColonyMovement%in%c("Out","In"),])+
  theme_classic(base_size = 16)+labs(color="TripNum")
```


## trip segmentation

Another common task in processing tracking data is to segment a trip into a outbound, middle, and inbound leg.  This can be done tediously by anotating each trip by hand or you can try using the `InOutPoints` function in `trakR`. 

```{r}
tracks_w_trips$INOut<-InOutPoints(tracks=tracks_w_trips,
                          CaptureID="CaptureID",
                          DateTime="DateTime",
                          TripID="TripNum",
                          dist2colony="Dist2Colony",lag=3,
                          nPointsToSmooth =10,minDist2Col = 10,
                          Lon = "Longitude",Lat="Latitude",
                          Plot = T,pdfName = "inout_plots.pdf")

ggplot(tracks_w_trips[tracks_w_trips$TripNum%in%c(5,7),],
       aes(Longitude,Latitude,col=factor(TripNum)))+
  geom_path(aes(lty=INOut),size=1)+
  scale_linetype_manual(values = c(1,2,3))+
  facet_wrap(~TripNum)+
  theme_classic(base_size = 16)+labs(color="TripNum")
```