---
title: "Split a track by returns to a central point"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Split a track by returns to a central point}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggplot2)
library(trakR)
```

# Load tracks 
This should be one file for all the tracks or create a data.frame with all the tracks.
The function `MakeTrip` is set up for processing multiple animals at once.
 
For this tutorial we will use a single bird in the included `tracks` dataset.
```{r load data}
data("tracks")

head(tracks)
```


The `MakeTrip` function requiers a column with distances to the colony for each point.  The `Dist2Colony` function will calculate these distances.

```{r dist2col}
tracks$Dist2Col<-trakR::Dist2Colony(tracks = tracks, 
                                    dataLat = "Latitude",
                                    dataLon = "Longitude",
                                    ColonyLat = 56.60329,
                                    ColonyLong = -169.6760)
```

## Split Trips

The `makeTrip` function finds the points where a animal moves some treshold distance `DistCutOff` away from the colony and the points where the animal returns within that threshold and annotates the data.frame with two new column, one indicating those "In" and "Out" points and another with trip numbers.  
```{r make trips}

tracks_w_trips<-trakR::MakeTrip(tracks = tracks,
                                ID = "CaptureID",
                                DistCutOff = 0.1,
                                Dist2Colony = "Dist2Col",
                                NumLocCut = 3)
head(tracks_w_trips)

```

You can see that the track was split into multiple trips
```{r plot,fig.asp=1}

# Plot a bird to check
ggplot(tracks_w_trips,aes(Longitude,Latitude,col=factor(TripNum)))+
  geom_path(size=.7)+
  geom_point(data=tracks_w_trips[tracks_w_trips$ColonyMovement%in%c("Out","In"),])+
  theme_classic(base_size = 16)+labs(color="TripNum")
```

## trip segmentation

Another common task in processing tracking data is to segment a trip into a outbound, middle, and inbound leg.  This can be done tediously by anotating each trip by hand or you can try using the `InOutPoints` function in `trakR`. 

```{r}
tracks_w_trips$INOut<-InOutPoints(tracks=tracks_w_trips,
                          CaptureID="CaptureID",
                          DateTime="DateTime",
                          TripID="TripNum",
                          dist2colony="Dist2Colony",lag=3,
                          nPointsToSmooth =10,minDist2Col = 10,
                          Lon = "Longitude",Lat="Latitude",
                          Plot = T,pdfName = "inout_plots.pdf")

ggplot(tracks_w_trips[tracks_w_trips$TripNum%in%c(5,7),],
       aes(Longitude,Latitude,col=factor(TripNum)))+
  geom_path(aes(lty=INOut),size=1)+
  scale_linetype_manual(values = c(1,2,3))+
  facet_wrap(~TripNum)+
  theme_classic(base_size = 16)+labs(color="TripNum")
```

## Extract Depth along a track

Sometimes it is helpful to attach sea floor depth or terrestial elevation to a location or series of locations (track).
The ETOPO1 dataset can be accessed with the `marmap` package witht he get.NOAA.bathy function.  This is a wrapper around that function that downloads the grid of elevation/depth data from the NOAA servers and then extracts the value for each cell that a point over lays and returns a vector of the same length as the data.

```{r}
min(tracks_w_trips$Longitude)
max(tracks_w_trips$Longitude)
min(tracks_w_trips$Latitude)
max(tracks_w_trips$Latitude)
tracks_w_trips$depth<-DepthETOPO(tracks,
                     dataLat="Latitude",
                     dataLon="Longitude",
                     left=min(tracks_w_trips$Longitude),
                     right= max(tracks_w_trips$Longitude),
                     top=min(tracks_w_trips$Latitude),
                     bottom=max(tracks_w_trips$Latitude),resolution = 1, keep=F)

# plot
ggplot(tracks_w_trips[tracks_w_trips$TripNum%in%c(5,7),],
       aes(Longitude,Latitude,col=depth))+
  geom_point(size=1)+
  scale_color_viridis_c()+
  facet_wrap(~TripNum)+
  theme_classic(base_size = 16)+labs(color="TripNum")
```

## Label events

Sometimes it is useful to number events consecutively such as forageing
events for a seabird.  often times these events have multiple rows per
event/bout.  This function numbers these evens consecutively by bird and by
trip nested within bird.
The function `label_event` is writen just for this purpose.  In this example we will label foraging events (in this case when birds drop below 5 km/h) with a consecutive number. 
It returns a data.frame with two new columns `event_trip` and `even_bird`

```{r}
tracks_w_trips$dtime<-InterpointTime(tracks_w_trips,ID = 'CaptureID',DateTime ='DateTime' )
tracks_w_trips$ddist<-InterpointDist(tracks_w_trips,ID = 'CaptureID',lat = 'Latitude',lon = 'Longitude' )
tracks_w_trips$ddist<-tracks_w_trips$ddist 
tracks_w_trips$speed<-Speed(Dist = tracks_w_trips$ddist,Time = tracks_w_trips$dtime)
head(tracks_w_trips)
hist(tracks_w_trips$speed,breaks=300)
tracks_w_trips$foraging<-tracks_w_trips$speed<5&tracks_w_trips$Dist2Col>10
tracks_w_trips<-label_events(dat=tracks_w_trips[!is.na(tracks_w_trips$foraging),],
                                                               BirdID = 'CaptureID',
                                                               TripID = 'TripNum',
                                                               eventCol = 'foraging')
head(tracks_w_trips)

ggplot(tracks_w_trips[tracks_w_trips$TripNum%in%c(5,7),],
       aes(Longitude,Latitude,col=factor(event_bird)))+
  geom_point(size=1)+
  scale_color_viridis_d()+
  facet_wrap(~TripNum)+
  theme_classic(base_size = 16)+labs(color="TripNum")
```
